---
title: "Supplementary materials for paper: French trade axe morphology differs by makers' mark: A preliminary analysis"
author: "Robert Z. Selden, Jr."
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: github_document
bibliography: book.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Load packages

```{r load.packages, echo = TRUE}
# devtools::install_github("geomorphR/geomorph", ref = "Stable", build_vignettes = TRUE)
library(here)
library(StereoMorph)
library(geomorph)
library(tidyverse)
library(wesanderson)
```

```{r load.data, include=FALSE}
# read data and define number of sLMs
shapes <- readShapes("./shapes")
coords<- readland.shapes(shapes, nCurvePts = c(5,10,10,10,3,5,5))

# read qualitative data
qdata <- read.csv("qdata.csv", header = TRUE, row.names = 1)
```

## Generalized Procrustes Analysis

Landmark data were aligned to a global coordinate system [@RN11622;@RN11623;@RN11563], achieved through generalized Procrustes superimposition [@RN478] performed in R 4.0.2 [@R] using the `geomorph` library v. 3.3.1 [@RN11530;@RN1774]. Procrustes superimposition translates, scales, and rotates the coordinate data to allow for comparisons among objects [@RN11564;@RN478]. The `geomorph` package uses a partial Procrustes superimposition that projects the aligned specimens into tangent space subsequent to alignment in preparation for the use of multivariate methods that assume linear space [@RN1646;@RN11563]. 

```{r gpa, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# gpa ----
Y.gpa <- gpagen(coords, print.progress = FALSE)
plot(Y.gpa)

# geomorph data frame ----
gdf <- geomorph.data.frame(shape = Y.gpa$coords, 
                           size = Y.gpa$Csize,
                           mark = qdata$mark)

# add centroid size to qdata ----
qdata$csz <- Y.gpa$Csize

# print qdata
knitr::kable(qdata, align = "lccc", caption = "Attributes included in qdata.")
```

### Variation in French trade axe size by mark

```{r box.attr, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# attributes for boxplots ----
csz <- qdata$csz # centroid size
cask <- qdata$cask # cask
mark <- qdata$mark  # mark

# boxplot of axe centroid size by mark ----
csz.mark <- ggplot(qdata, aes(x = mark, y = csz, color = mark)) + 
  geom_boxplot() +
  geom_dotplot(binaxis = 'y', stackdir = 'center', dotsize = 0.3) +
  scale_colour_manual(values = wes_palette("Moonrise2")) +
  theme(legend.position = "none") +
  labs(x = 'Mark', y = 'Centroid Size')
# render plot
csz.mark
```

## Principal Components Analysis

Principal components analysis [@RN1746] was used to visualise shape variation among the axes The shape changes described by each principal axis are commonly visualized using thin-plate spline warping of a reference 3D mesh [@RN1731;@RN479]. 

```{r pca, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# principal components analysis
pca<-gm.prcomp(Y.gpa$coords)
summary(pca)

# set plot parameters to plot by mark
pch.gps <- c(15,17)[as.factor(mark)]
col.gps <- wes_palette("Moonrise2")[as.factor(mark)]
col.hull <- c("#798E87","#C27D38")

# plot pca by mark
pc.plot1 <- plot(pca, 
                 asp = 1,
                 pch = pch.gps,
                 col = col.gps)
shapeHulls(pc.plot1, 
           groups = mark,
           group.cols = col.hull)
```

## Test hypothesis

_Hypothesis: There are morphological differences in French trade axes that bear an asterisk or DG makers' marks._

```{r def.mod.1, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
## size as a function of mark ----
fit.size.mark <- procD.lm(size ~ mark, 
                          data = gdf, 
                          print.progress = FALSE, 
                          iter = 9999)

# differences in size by mark?
anova(fit.size.mark)

## shape as a function of mark ----
fit.shape.mark <- procD.lm(shape ~ mark, 
                           data = gdf, 
                           print.progress = FALSE, 
                           iter = 9999)

## differences in shape by mark? ----
anova(fit.shape.mark)
```

### Mean shapes

```{r def.mod.2, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# mean shapes ----
new.coords<-coords.subset(A = Y.gpa$coords, 
                          group = qdata$mark)
names(new.coords)

# group shape means
mean <- lapply(new.coords, mshape)

# plot mean shapes
plot(mean$asterisk) # mean shape of axe with asterisk mark
plot(mean$DG) # mean shape of axe with DG mark

# comparison plot of asterisk (gray) vs DG (linear vector)
plotRefToTarget(mean$asterisk, 
                mean$DG, 
                method = "vector",
                mag = 2)
```

## References cited